Liveness-Based Garbage Collection for Lazy Languages

### 2015-11-28 paper/paper.tex
latest - 27th

  %% We  consider  a  round of garbage  collection  followed  by  a
  %% transition as a step.  We show by induction that, for
  %% programs  without function  calls, %% starting  from the
  %% %% initial  state  that has $e_{\main}$  as  the  
  %% %% expression to  be evaluated,
  %% any
  %% transition of $n$ steps occurs without a \bang.

@@

Note that our proofs will be for a single round of minefield execution
i.e.  the evaluation of $(\mainpgm)$ (or $e_{\mainpgm}$, for call-less
programs) to  its WHNF driven  by the printing mechanism.   With minor
variations,  the   proof  will  also  be   applicable  for  subsequent
evaluations initiated by $\print$.
%% We  show next that  the minefield execution of  a call-less
%% program cannot go \bang.

@@

For the inductive step, we  shall show that none of the
minefield steps that  involves dereferencing results in
a   \bang.   These  are   the   steps   which  have   a
$\heap(\rho(...))$ in the premise. %%  Recall that a minefield
%% step   consists   of   a   GC  step   followed   by   a
%% $\rightsquigarrow$. 
Now a  $\rightsquigarrow$ step can  go
\bang\ because it dereferences a $\bot$ inserted by the
immediately  preceding  GC  or  the GC  of  an  earlier
minefield step.  However the  demand $\sigma'$
  on basis of which an earlier GC would have inserted a
  $\bot$,  would  have   included  the  current  demand
  $\sigma$.  Thus,  if the  earlier GC had  inserted a
$\bot$, so would the current  GC.  Thus it is enough to
consider the $\rightsquigarrow$  along with the current
GC and show that it does not lead to a \bang.

@@

In  summary, the  equations  generated during  liveness
analysis are:
\begin{enumerate}
\item   For  each   function  $\mathit{f}$,   equations
  defining \Df{f}{i} for use by \Lfonly$_{\mathit f}$.
\item For each function $\mathit{f}$, an equation defining the summary
  demand $\sigma_{\mathit f}$ on $e_f$.
\item   For  each   function  $\mathit{f}$   (including
  $\mainpgm$  for  $e_\mainpgm$) an  equation  defining
  liveness at each GC point of $e_{\mathit f}$.
\end{enumerate}
\subsection{Solving liveness equations---the grammar
interpretation}\label{sec:grammar-formulation}      The
equations  above   can  now  be  re-interpreted   as  a
context-free    grammar   (CFG)    on   the    alphabet
$\lbrace\acar,  \acdr,  \bcar,  \bcdr,  \clazy\rbrace$.
Let \var{$X$}  denote the  non-terminal for  a variable
$X$  occurring on  the LHS  of the  equations generated
from  the  analysis.  We  can  think  of the  resulting
productions as being  associated with several grammars,
one for each non-terminal \var{\Lanv{i}{x}} regarded as
a  start  symbol.   As  an  example,  the  grammar  for
\var{\Lanv{1}{\xl}}     comprises     the     following
productions:
\begin{eqnarray*}
  \var{\Lanv{3}{\xl}}  &\rightarrow& 
  \clazy \mid \var{\Df {\Sum}{1}}{\clazy\var{\sigma_{\length}}}  \\
  & & \mid
  \clazy\var{\Df{\length}{1}}\clazy \mid \acar\var{\Df{\fun}{2}}\bcdr\var{\sigma_{\mathit{\fun}}} \\
  \var{\Df{\length}{1}} &\rightarrow& \clazy \mid
  \acdr\var{\Df{\length}{1}}\clazy
       \mid \clazy\var{\Df{\length}{1}}{\clazy}\\
\langle {\sigma_{\length}} \rangle
&\rightarrow&
\clazy  \mid \bcar\langle{\sigma_{\Sum}}\rangle \\
 \var{\Df{\fun}{2}} &\rightarrow& \var{\Df{\makelist}{1}}\clazy \mid \var{\Df{\makelist}{1}}\var{\Df{\Sum}{1}}\clazy \\ 
& & \mid \var{\Df{\makelist}{1}}\acar\var{\Df{\fun}{2}}\bcdr\var{\sigma_{\mathit{\fun}}} \\
\langle\sigma_{\mathit{\fun}} \rangle &\rightarrow& \sigma_{all} \mid \bcdr\var{\sigma_{\mathit{\fun}}} \\
\var{\Df{\Sum}{1}} &\rightarrow& \clazy \mid \acar\clazy \mid \acdr\var{\Df{\Sum}{1}}\clazy
%% \var{\Lanv{3}{\py}} &\rightarrow& \clazy \mid \bcar\var{\sigma_{\mathit{\fun}}}
\end{eqnarray*}
Other  equations  can   be  converted  similarly.   The
language   generated   by  \var{\Lanv{i}{x}},   denoted
$\mathscr{L}(\var{\Lanv{i}{x}})$,   is    the   desired
solution  of  \Lanv{i}{x}.    However,  note  that  the
strings  in the  language  are  over alphabet  $\lbrace
\acar, \acdr,\bcar, \bcdr, \clazy \rbrace$, but we want
     {\em  forward} access  paths only,  i. e.,  access
     paths  over  the  alphabet $\lbrace  \acar,  \acdr
     \rbrace$.  In other  words,  the decision  problem
     that   we  are   interested   in  during   garbage
     collection at a program point $\pi_i$ is:
\begin{quote}
Let $x.\alpha$ be a forward access path consisting only
of edges  \acar\ and \acdr\  (but not \bcar,  \bcdr\ or
\clazy).       Let      $\mathscr{L}(\var{\Lanv{i}{x}})
\stackrel{*}{\hookrightarrow}  \sigma$, where  $\sigma$
consists  of  forward  access  paths  only.  Then  does
$\alpha \in \sigma$?
\end{quote}

@@

From Sections \ref{sec:liveness}  and \ref{sec:computing}, it is clear
that the results of liveness analysis of any program can be modeled by
the kind  of grammar described above. The  following proposition shows
that the converse also holds.
\begin{proposition}
Given    a     grammar    $G$    of    the     form    described    in
Definition~\ref{def:specialgrammar},  it is  possible  to construct  a
program $p$  with program points  $\pi_i$ and variables $x$  such that
the liveness  analysis of $p$  is same as  $G$ except for a  change in
non-terminal names which now have the form \var{\Lanv{i}{x}}.
\end{proposition}
We now show that the decision  problem that is required to be answered
at       garbage      collection       time       is      undecidable.
\newcommand{\state}{\ensuremath{\mathsf{S}}}
\newcommand{\nont}[2]{\ensuremath{\mathsf{S}_{#1}^{#2}}}  
\begin{lemma}
Consider    a    grammar    $G$    of   the    kind    described    in
Definition~\ref{def:specialgrammar} and a forward access path $\alpha$
consisting  of symbols \acar\  and \acdr\  only. The  decision problem
$\alpha \in \mathscr{L}(G)$ is undecidable.
\end{lemma}

@@

We shall denote the configuration of the Turing machine
as $w_l(\state,c)w_r$, where $w_l$ is the string to the
left of the head, $w_r$ is the string to the right, $c$
is the symbol under the head and \state\ is the current
state of  the machine.   For each combination  of state
and symbol  $(\state,c)$, the grammar will  contain the
non-terminal \nont{}{c}. We shall synchronize each move
of the  machine to  a derivation  step using  a context
free production, followed, if possible, by a derivation
step using either  $\bcar\acar \rightarrow \epsilon$ or
$\bcdr\acdr   \rightarrow    \epsilon$.    After   each
synchronization,  we  shall   establish  the  following
invariant  relation between  the machine  configuration
and the sentential form:

\begin{quote}
  If    the   configuration    of   the    machine   is
  $w_l(\state,c)w_r$, then the  sentential form will be
  $\overline{w}_l\nont{}{c}\,w_r$,                where
  $\overline{w}_l$ is  the same as $w_l$  but with each
  symbol $d$ in $w_l$ replaced by $\overline{d}$.
\end{quote}

Assume that  the TM starts  in a  state $S_\mathit{init}$ with  a tape
$cw$ and  the head positioned on  the symbol $c$. Then  the sentential
form    corresponding     to    the    initial     configuration    is
$\nont{\mathit{init}}{c}w$ (we  can assume that there  is a production
$\nont{}{} \rightarrow  \nont{\mathit{init}}{c}w$, where  \nont{}{} is
the start symbol of the  grammar). Further correspondences between the
Turing machine moves and the grammar productions are as follows:

\begin{enumerate}
\item For each transition  $(S_i, c) \rightarrow (S_j,c',L)$, there are
  two  productions $\nont{i}{c}  \rightarrow  \acar \nont{j}{\acar}c'$
  and $\nont{i}{c} \rightarrow \acdr \nont{j}{\acdr}c'$.
\item For each  transition $(S_i, c) \rightarrow (S_j,c',R)$, there
  are two productions
  $\nont{i}{c} \rightarrow c \nont{j}{\acar}\bcar$ and $\nont{i}{c}
  \rightarrow c \nont{j}{\acdr}\bcdr$. 
\end{enumerate}
The idea  behind the  productions is explained  with an
example:  Assume that  the current  sentential form  is
\bcar\bcdr\nont{i}{\acar}\acar\acar.   Assume that  the
machine  has  a   transition  $(S_i,\acar)  \rightarrow
(S_j,\acdr, L)$.  Since the next corresponding  step in
the  derivation  has  to  be  done  without  any  prior
knowledge of whether the symbol to the left of the tape
is a \acar\  or a \acdr, two  productions are provided,
and  the  invariant  will  be maintained  only  if  the
production         $\nont{i}{\acar}         \rightarrow
\acdr\nont{j}{\acdr}\acdr$ is chosen  for the next step
in  the  derivation.    This  gives  the  configuration
\bcar\bcdr\acdr\nont{i}{\acdr}\acdr\acar\acar.
Simplification   with    the   production   $\bcdr\acdr
\rightarrow               \epsilon$              yields
\bcar\nont{i}{\acdr}\acdr\acar\acar,    which   exactly
corresponds  to   the  changed  configuration   of  the
machine.  Notice  carefully that a wrong  choice breaks
the invariant  and it cannot be  recovered subsequently
by any choice of productions.

After  the Turing machine  has halted,  there are  further ``cleanup''
derivations  that derive  $\epsilon$ only  if the  invariant  has been
maintained  so far.  For every  symbol  $c$, we introduce  a  non-terminal
$\nont{\mathit{final}}{c}$ where $\nont{\mathit{final}}{}$ is the
final state of the Turing machine. We add
productions $\nont{\mathit{final}}{c} \rightarrow \acar
\nont{\mathit{final}}{c}$ and $\nont{\mathit{final}}{c} \rightarrow \acdr
\nont{\mathit{final}}{c}$ for cleaning up the \bcar\ and \bcdr\ symbols on
the left of the head and $\nont{\mathit{final}}{c} \rightarrow 
\nont{\mathit{final}}{c}\bcar$ and $\nont{\mathit{final}}{c} \rightarrow 
\nont{\mathit{final}}{c}\bcdr$ for cleaning up \acar\ and \acdr\ on the
right of the tape head. This completes the reduction.
\end{proof}

@@

\subsubsection{From CFGs to GC-ready DFA}
\label{sec:NFA-approx}
 We use  the algorithm by Mohri  and Nederhof~\cite{mohri00regular} to
 approximate a CFG by a {\em strongly regular\/} grammar.  For
 example, the grammar
 fragment  for  the  non-terminal  $\var{\Df{\length}{1}}$  after  the
 Mohri-Nederhof transformation is:
 \begin{eqnarray*}
   \var{\Df{\length}{1}} &\rightarrow& \clazy\var{\Df{\length}{1'}} \mid
   \acdr\var{\Df{\length}{1}}
   \mid \clazy\var{\Df{\length}{1}}\\
   \var{\Df{\length}{1'}} &\rightarrow& \clazy\var{\Df{\length}{1'}}
   \mid \epsilon
 \end{eqnarray*}

@@

%\subsection{Garbage collection for references on print stack}
The evaluation of  the top-level expression in a program  is driven by
the \print\  function.  Printing of  atomic values is simple  and does
not  trigger a  garbage  collection.   In case  of  \CONS\ cells,  the
\CAR\ is  first printed followed  by \CDR.   In a lazy  language, both
\CAR\ and \CDR\  could be closures requiring evaluation,  and this may
trigger  more garbage  collections.  When  this happens.   we have  to
consider any  references that  might be  on the  print stack  and copy
them.  We extend liveness-based  garbage  collection  to
the function \print.

>>>>

%% We consider a round of  garbage collection followed by a transition
%% as  a  step.  We  show  by  induction  that, for  programs  without
%% function  calls, %%  starting from  the %%  initial state  that has
%% $e_{\main}$ as the %% expression to be evaluated, any transition of
%% $n$ steps occurs without a \bang.

@@

Note that our proofs will be for a single round of minefield execution
i.e.  the evaluation of $(\mainpgm)$ (or $e_{\mainpgm}$, for call-less
programs) to  its WHNF driven  by the printing mechanism.   With minor
variations,  the   proof  will  also  be   applicable  for  subsequent
evaluations initiated by $\print$.
%% We show  next that the  minefield execution of a  call-less program
%% cannot go \bang.

@@

For the inductive step, we shall show that none of the minefield steps
that involves dereferencing  results in a \bang.  These  are the steps
which  have a  $\heap(\rho(...))$ in  the  premise. 
Now a  $\rightsquigarrow$ step can  go
\bang\ because it dereferences a $\bot$ inserted by the
immediately  preceding  GC  or  the GC  of  an  earlier
minefield step.  However the  demand $\sigma'$
  on basis of which an earlier GC would have inserted a
  $\bot$,  would  have   included  the  current  demand
  $\sigma$.  Thus,  if the  earlier GC had  inserted a
$\bot$, so would the current  GC.  Thus it is enough to
consider the $\rightsquigarrow$  along with the current
GC and show that it does not lead to a \bang.

@@

In  summary, the  equations  generated during  liveness
analysis are:
\begin{enumerate}
\item   For  each   function  $\mathit{f}$,   equations
  defining \Df{f}{i} for use by \Lfonly$_{\mathit f}$.
\item For each function $\mathit{f}$, an equation defining the summary
  demand $\sigma_{\mathit f}$ on $e_f$.
\item   For  each   function  $\mathit{f}$   (including
  $\mainpgm$  for  $e_\mainpgm$) an  equation  defining
  liveness at each GC point of $e_{\mathit f}$.
\end{enumerate}
\subsection{Solving liveness equations---the grammar
interpretation}\label{sec:grammar-formulation}      The
equations  above   can  now  be  re-interpreted   as  a
context-free    grammar   (CFG)    on   the    alphabet
$\lbrace\acar,  \acdr,  \bcar,  \bcdr,  \clazy\rbrace$.
Let \var{$X$}  denote the  non-terminal for  a variable
$X$  occurring on  the LHS  of the  equations generated
from  the  analysis.  We  can  think  of the  resulting
productions as being  associated with several grammars,
one for each non-terminal \var{\Lanv{i}{x}} regarded as
a  start  symbol.   As  an  example,  the  grammar  for
\var{\Lanv{1}{\xl}}     comprises     the     following
productions:
\begin{eqnarray*}
  \var{\Lanv{3}{\xl}}  &\rightarrow& 
  \clazy \mid \var{\Df {\Sum}{1}}{\clazy\var{\sigma_{\length}}}  \\
  & & \mid
  \clazy\var{\Df{\length}{1}}\clazy \mid \acar\var{\Df{\fun}{2}}\bcdr\var{\sigma_{\mathit{\fun}}} \\
  \var{\Df{\length}{1}} &\rightarrow& \clazy \mid
  \acdr\var{\Df{\length}{1}}\clazy
       \mid \clazy\var{\Df{\length}{1}}{\clazy}\\
\langle {\sigma_{\length}} \rangle
&\rightarrow&
\clazy  \mid \bcar\langle{\sigma_{\Sum}}\rangle \\
 \var{\Df{\fun}{2}} &\rightarrow& \var{\Df{\makelist}{1}}\clazy \mid \var{\Df{\makelist}{1}}\var{\Df{\Sum}{1}}\clazy \\ 
& & \mid \var{\Df{\makelist}{1}}\acar\var{\Df{\fun}{2}}\bcdr\var{\sigma_{\mathit{\fun}}} \\
\langle\sigma_{\mathit{\fun}} \rangle &\rightarrow& \sigma_{all} \mid \bcdr\var{\sigma_{\mathit{\fun}}} \\
\var{\Df{\Sum}{1}} &\rightarrow& \clazy \mid \acar\clazy \mid \acdr\var{\Df{\Sum}{1}}\clazy
%% \var{\Lanv{3}{\py}} &\rightarrow& \clazy \mid \bcar\var{\sigma_{\mathit{\fun}}}
\end{eqnarray*}
Other  equations  can   be  converted  similarly.   The
language   generated   by  \var{\Lanv{i}{x}},   denoted
$\mathscr{L}(\var{\Lanv{i}{x}})$,   is    the   desired
solution  of  \Lanv{i}{x}.

\warning{
\hrule
\medskip
Fixing from here
\medskip
\hrule
}

However, recall that the decision problem that we are interested in
during garbage collection is the following:
\begin{quote}
Let $x.\alpha$ be a forward access path consisting only
of edges  \acar\ and \acdr\  (but not \bcar,  \bcdr\ or
\clazy).       Let      $\mathscr{L}(\var{\Lanv{i}{x}})
\stackrel{*}{\hookrightarrow}  \sigma$, where  $\sigma$
consists  of  forward  access  paths  only.  Then  does
$\alpha \in \sigma$?
\end{quote}

@@

%% From Sections \ref{sec:liveness}  and \ref{sec:computing}, it is clear
%% that the results of liveness analysis of any program can be modeled by
%% the kind  of grammar described above. The  following proposition shows
%% that the converse also holds.
%% \begin{proposition}
%% Given    a     grammar    $G$    of    the     form    described    in
%% Definition~\ref{def:specialgrammar},  it is  possible  to construct  a
%% program $p$  with program points  $\pi_i$ and variables $x$  such that
%% the liveness  analysis of $p$  is same as  $G$ except for a  change in
%% non-terminal names which now have the form \var{\Lanv{i}{x}}.
%% \end{proposition}
We first show that membership problem of the class of grammars in
Definition~\ref{def:specialgrammar} is undecidable. However note that
the grammar may not be necessarily generated from liveness analysis.
\newcommand{\state}{\ensuremath{\mathsf{S}}}
\newcommand{\nont}[2]{\ensuremath{\mathsf{S}_{#1}^{#2}}}  
\begin{lemma}\label{lemma:grammar-undecidable}
Consider a grammar    $G$    of   the    kind    described    in
Definition~\ref{def:specialgrammar}  and a forward access path $\alpha$
consisting  of symbols \acar\  and \acdr\  only. The membership problem
$\alpha\$ \in \mathscr{L}(G)$ is undecidable.
\end{lemma}

@@

We shall represent a Turing machine (TM) configuration 
as $w_l(\state,c)w_r$, where $w_l$ is the string to the
left of the head, $w_r$ is the string to the right, $c$
is the symbol under the head and \state\ is the current
state of  the TM.   For each combination  of state
and symbol  $(\state,c)$, the grammar will  contain the
non-terminal \nont{}{c}. We shall synchronize each move
of the TM to  a derivation  step using  a context
free production, followed, if possible, by a derivation
step using either  $\bcar\acar \rightarrow \epsilon$ or
$\bcdr\acdr   \rightarrow    \epsilon$.    After   each
synchronization,  we  shall   establish  the  following
invariant  relation between  the TM  configuration
and the sentential form:

\begin{quote}
  If    the   configuration    of   the    TM   is
  $w_l(\state,c)w_r$, then the  sentential form will be
  $\overline{w}_l\nont{}{c}\,w_r$,                where
  $\overline{w}_l$ is  the same as $w_l$  but with each
  symbol $d$ in $w_l$ replaced by $\overline{d}$.
\end{quote}

Assume that  the TM starts  in a  state $S_\mathit{init}$ with  a tape
$cw$ and  the head positioned on  the symbol $c$. Then  the sentential
form    corresponding     to    the    initial     configuration    is
$\nont{\mathit{init}}{c}w\$$ (we  can assume that there  is a production
$\nont{}{} \rightarrow  \nont{\mathit{init}}{c}w\$$, where  \nont{}{} is
the start symbol of the  grammar). Further correspondences between the
TM moves and the grammar productions are as follows:

\begin{enumerate}
\item For each transition  $(S_i, c) \rightarrow (S_j,c',L)$, there are
  two  productions $\nont{i}{c}  \rightarrow  \acar \nont{j}{\acar}c'$
  and $\nont{i}{c} \rightarrow \acdr \nont{j}{\acdr}c'$.
\item For each  transition $(S_i, c) \rightarrow (S_j,c',R)$, there
  are two productions
  $\nont{i}{c} \rightarrow c \nont{j}{\acar}\bcar$ and $\nont{i}{c}
  \rightarrow c \nont{j}{\acdr}\bcdr$. 
\end{enumerate}
The idea  behind the  productions is explained  with an
example:  Assume that  the current  sentential form  is
\bcar\bcdr\nont{i}{\acar}\acar\acar. Also  assume that  the
TM  has  a   transition  $(S_i,\acar)  \rightarrow
(S_j,\acdr, L)$.  Since the next corresponding  step in
the  derivation  has  to  be  done  without  any  prior
knowledge of whether the symbol to the left of the tape
is a \acar\  or a \acdr, two  productions are provided,
and  the  invariant  will  be maintained  only  if  the
production         $\nont{i}{\acar}         \rightarrow
\acdr\nont{j}{\acdr}\acdr$ is chosen  for the next step
in  the  derivation.    This  gives  the  configuration
\bcar\bcdr\acdr\nont{i}{\acdr}\acdr\acar\acar.
Simplification   with    the   production   $\bcdr\acdr
\rightarrow               \epsilon$              yields
\bcar\nont{i}{\acdr}\acdr\acar\acar,    which   exactly
corresponds  to   the  changed  configuration   of  the
TM.  Notice  carefully that a wrong  choice breaks
the invariant  and it cannot be  recovered subsequently
by any choice of productions.

After  the TM  has halted,  there are  further ``cleanup''
derivations  that derive  $\epsilon$ only  if the  invariant  has been
maintained  so far.  For every  symbol  $c$, we introduce  a  non-terminal
$\nont{\mathit{final}}{c}$ where $\nont{\mathit{final}}{}$ is the
final state of the TM. We add
productions $\nont{\mathit{final}}{c} \rightarrow \acar
\nont{\mathit{final}}{c}$ and $\nont{\mathit{final}}{c} \rightarrow \acdr
\nont{\mathit{final}}{c}$ for cleaning up the \bcar\ and \bcdr\ symbols on
the left of the head and $\nont{\mathit{final}}{c} \rightarrow 
\nont{\mathit{final}}{c}\bcar$ and $\nont{\mathit{final}}{c} \rightarrow 
\nont{\mathit{final}}{c}\bcdr$ for cleaning up \acar\ and \acdr\ on the
right of the tape head. Finally the production $\clazy\$ \rightarrow
\epsilon$ is used to derive $\epsilon$.This completes the reduction.
\end{proof}

We now show that the proof can be replayed for the class of grammars
generated from liveness analysis of programs.

\begin{lemma}
Consider a grammar    $G$    of   the    kind    described    in
Definition~\ref{def:specialgrammar} that is generated by liveness analysis
of a program and a forward access path $\alpha$. The membership problem
$\alpha\$ \in \mathscr{L}(G)$ is also undecidable.
\end{lemma}

Given a Turing machine and an input string, we now give program fragments whose liveness
analysis generate the same productions as in
Lemma~\ref{lemma:grammar-undecidable}, but with  extra productions
that necessarily contain the \clazy\ symbol. For instance,
corresponding to the production $\nont{i}{c} \rightarrow c
\nont{j}{\acar}\bcar$, with $c$ as \acar\ for concreteness, we have
the program (define ($\nont{i}{c}$ a) (let b <- (car a) in (let c <-
($\nont{j}{\acar}$ b) in (let d <- 5 in (let e <- (cons c d) in (let f
<- (null? e) in return f)))))). The reader can check that this
produces the grammar fragment ...

It is obvious that if the TM halts on the input, then the grammar will
produce the $\epsilon$ string. However, if the TM does not halt on the
input, then the productions containg

@@

\subsubsection{From CFGs to GC-ready DFAs}
\label{sec:NFA-approx}
 We use  the algorithm by Mohri  and Nederhof~\cite{mohri00regular} to
 approximate a CFG by a {\em strongly regular\/} grammar.  For
 example, the grammar
 fragment  for  the  non-terminal  $\var{\Df{\length}{1}}$  after  the
 Mohri-Nederhof transformation is:
 \begin{eqnarray*}
   \var{\Df{\length}{1}} &\rightarrow& \clazy\var{\Df{\length}{1'}} \mid
   \acdr\var{\Df{\length}{1}}
   \mid \clazy\var{\Df{\length}{1}}\\
   \var{\Df{\length}{1'}} &\rightarrow& \clazy\var{\Df{\length}{1'}}
   \mid \epsilon
 \end{eqnarray*}

@@

%\subsection{Garbage collection for references on print stack}
The evaluation of  the top-level expression in a program  is driven by
a printing function.  
%% Printing of  atomic values is simple  and does
%% not  trigger a  garbage  collection.   In case  of  \CONS\ cells,  the
%% \CAR\ is  first printed followed  by \CDR.   In a lazy  language, both
%% \CAR\ and \CDR\  could be closures requiring evaluation,  and this may
%% trigger  more garbage  collections.  When  this happens.   we have  to
%% consider any  references that  might be  on the  print stack  and copy
%% them.
We extend liveness-based  garbage  collection  to
this function.
